Probability Aggregation: Dynamic Hierarchical Modeling of Expert Beliefs in Geopolitical Events

### 2013-08-23 paper.tex
Edits

\section{Introduction}
\label{intro}
Individual experts can differ radically from one another in their abilities to assess probabilities of future events. The quality of a probability assessment is typically evaluated in regards to an important attribute called \textit{calibration}. An expert is said to be well-calibrated if his probabilities agree with the long-term frequency of the events. This means that, for instance, the proportion of events that actually occurred is 60\% for all those events for which he assessed a probability of 0.60. Even though several experiments have been conducted to show that experts are generally poorly calibrated (see, e.g., \citet{cooke1991experts, shlyakhter1994quantifying}), the level of expertise has been found to be associated with calibration. In particular, \citet{wright1994coherence} argue that a higher self-reported expertise on the subject matter implies better calibration.

Unfortunately, calibration by itself does not guarantee usefulness. To see this, consider a relatively stationary process, such as rain in some geographic region, whose empirical frequency of occurrence in the last 10 years is 45\%. In this setting, an expert can assess a constant probability of 0.45 and be well-calibrated. This probability, however, is very near 0.5 and hence too uninformative to be useful in decision making. Compare this to the advice of  another expert whose typical probability assessment is near 0.0 or 1.0; that is, he is a \textit{sharp} assessor. If the expert is both sharp and well-calibrated, then he is able to forecast the behavior of the process both with high certainty and accuracy. From the decision-maker's perspective, such advice is highly useful as it involves very little risk. Therefore the goal in probability estimation should be to maximize sharpness subject to calibration (see, e.g., \citet{raftery2005using, Ranjan08}). This is a well-known goal that has led to a wide range of novel and insightful observations in probability forecasting literature.

@@

\subsection{Group-Level Expertise Bias}
\label{ExpertBias}
Recall from Section \ref{data} that the experts were asked to self-assess their expertise level (on a 1-to-5 scale with 1 = Not At All Expert to 5 = Extremely Expert) on  all of the 166 questions in our real-world dataset. Based on these self-assessments the experts were then divided into 5 expertise groups, and each of the expertise groups was assigned a separate multiplicative bias term. This section uses the \textit{Sample-Then-Calibrate}-procedure to explore the posterior distribution of these group-level bias terms. Figure \ref{Biases} presents the posterior distributions of the bias terms with side-by-side box plots. Since the distributions fall completely below the \textit{no-bias} reference-line at 1.0, the expertise groups are deemed under-confident.

The under-confidence, however, decreases as the level of the self-reported expertise increases. In order to understand this result, it is helpful to view the process of making a subjective probability assessment as Bayesian updating: A completely ignorant expert aiming to minimize a reasonable loss function, such as the Brier score, has no reason to give anything but 0.5 as his probability forecast. However, as soon as the forecaster gains some knowledge about the event, he produces an updated forecast that is a compromise between his initial forecast and the new information acquired. The updated forecast is therefore conservative and too close to 0.5 as long as the forecaster remains only partially informed about the event. If most forecasters fall somewhere on this spectrum between ignorance and full information, their average forecast tends to fall strictly between 0.5 and the most-informed probability forecast (see \citet{Baron} for more details). Since expertise is to a large extent determined by subject-matter knowledge, the level of under-confidence can be expected to decrease as a function of the group's level of self-reported expertise.

Overall, finding under-confidence is a rather surprising result given that many previous studies have been conducted to show that experts are likely to be over-confident. For instance,  \citet{lichtenstein1977calibration, morgan1992uncertainty, bier2004implications} summarize the results from numerous calibration studies, and conclude that experts are systematically over-conﬁdent about their probability assessments. Our result, however, is a statement about groups of experts and hence does not invalidate the possibility of the individual experts being overconfident. To make conclusions at the individual-level based on the group-level bias terms would result into a \textit{ecological inference fallacy} (see, e.g., \citet{lubinski1996seeing}).

@@

More specific group-level inference can be performed by computing probabilities based the estimated posterior distributions. For instance, the posterior probability that the most expert group is the least under-confident is approximately equal to $1.0$, and 
the posterior probability of a strictly decreasing level of under-confidence by expertise ranking is approximately 0.87. The latter probability is driven down by the inseparability of the two groups associated with the lowest levels of self-reported expertise. The fact that these groups are very similar suggests that the forecasters are poor at assessing how little they know about a subject that is strange to them. If these groups are combined into a single group, the posterior probability of a strictly decreasing level of under-confidence is approximately 1.0.

@@

\section{Discussion}
This paper began with an introduction of a rather unorthodox but nonetheless realistic probability forecasting setting, in which forecasts are made very infrequently by a heterogeneous group of experts over a period of time. Unfortunately, the resulting data is too sparse to be modeled efficiently with standard time-series methods.  In response to this lack of previous methodology, our work 
%expands the literature towards less-studied areas of  probability aggregation by
introduces an interpretable time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated crowd belief across time. The model is estimated in two-steps: The first step samples constrained versions of the model parameters via a Gibbs sampler that only makes use of standard distributions. The second step involves a one-dimensional optimization procedure that transforms the constrained parameter values into their unconstrained counterparts.

>>>>

\section{Introduction}
\label{intro}
Individual experts can differ radically from one another in their abilities to assess probabilities of future events. The quality of a probability assessment is typically evaluated in regards to an important attribute called \textit{calibration}. An expert is said to be well-calibrated if his probabilities match with the long-term frequencies of the events. This means that, for instance, the proportion of events that actually occurred is 60\% for all those events for which he assessed a probability of 0.60. Even though several experiments have been conducted to show that experts are generally poorly calibrated (see, e.g., \citet{cooke1991experts, shlyakhter1994quantifying}), expertise has been found to be associated with calibration. In particular, \citet{wright1994coherence} argue that a higher level of self-reported expertise on the subject matter implies better calibration.

Unfortunately, calibration by itself does not guarantee usefulness. To see this, consider a relatively stationary process, such as rain in some geographic region, whose empirical frequency of occurrence in the last 10 years is 45\%. In this setting an expert can assess a constant probability of 0.45 and be well-calibrated. This probability, however, is close to 0.5 and hence too uninformative to be very useful in decision making. For the sake of comparison, consider another expert whose typical probability assessment is near 0.0 or 1.0. In the forecasting literature such an expert is called \textit{sharp}. If the expert is not only sharp but also well-calibrated, then he is able to forecast the behavior of the process both with high certainty and accuracy. From the decision-maker's perspective, his advice is highly useful as it involves very little risk and much information. For these reasons, a common goal in probability estimation is to maximize sharpness subject to calibration (see, e.g., \citet{raftery2005using, Ranjan08}). This is a well-defined goal that has led to a wide range of novel and insightful observations in probability forecasting literature.

@@

\subsection{Group-Level Expertise Bias}
\label{ExpertBias}
Recall from Section \ref{data} that the experts were asked to self-assess their level of expertise (on a 1-to-5 scale with 1 = Not At All Expert to 5 = Extremely Expert) on any questions that they participated in. The self-reported expertise then divides the experts into 5  groups, with each group assigned a separate multiplicative bias term. This section uses the \textit{Sample-Then-Calibrate}-procedure to explore the posterior distribution of these multiplicative bias terms. Figure \ref{Biases} presents the posterior distributions of the bias terms with side-by-side box plots. Since the distributions fall completely below the \textit{no-bias} reference-line at 1.0, all the groups are deemed under-confident.

The under-confidence, however, decreases as the level of  expertise increases.  For instance, the posterior probability that the most expert group is the least under-confident is approximately equal to $1.0$, and the posterior probability of a strictly decreasing level of under-confidence is approximately 0.87. The latter probability is driven down by the inseparability of the two groups associated with the lowest levels of self-reported expertise. The fact that these groups are very similar suggests that the forecasters are poor at assessing how little they know about a subject that is strange to them. If these groups are combined into a single group, the posterior probability of a strictly decreasing level of under-confidence is approximately 1.0.

The decreasing trend in under-confidence can be reasoned by viewing the process of making a subjective probability  as Bayesian updating: At first a completely ignorant expert aiming to minimize a reasonable loss function, such as the Brier score, has no reason to give anything but 0.5 as his probability forecast. However, as soon as the expert gains some knowledge about the event, he produces an updated forecast that is a compromise between his initial forecast and the new information acquired. The updated forecast is therefore conservative and too close to 0.5 as long as the expert remains only partially informed about the event. If most experts fall somewhere on this spectrum between ignorance and full information, their average forecast tends to fall strictly between 0.5 and the most-informed probability forecast (see \citet{Baron} for more details). Since expertise is to a large extent determined by subject-matter knowledge, the level of under-confidence can be expected to decrease as a function of the group's level of self-reported expertise.

Finding under-confidence in all the groups is a rather surprising result given that many previous studies have been conducted to show that experts are likely to be over-confident. For instance,  \citet{lichtenstein1977calibration, morgan1992uncertainty, bier2004implications} summarize the results from numerous calibration studies, and conclude that experts are systematically over-conﬁdent about their probability assessments. Our result, however, is a statement about groups of experts and hence does not invalidate the possibility of the individual experts being overconfident. To make conclusions at the individual-level based on the group-level bias terms would be considered an \textit{ecological inference fallacy} (see, e.g., \citet{lubinski1996seeing}).

@@

%More specific group-level inference can be performed by computing probabilities based the estimated posterior distributions. For instance, the posterior probability that the most expert group is the least under-confident is approximately equal to $1.0$, and 
%the posterior probability of a strictly decreasing level of under-confidence by expertise ranking is approximately 0.87. The latter probability is driven down by the inseparability of the two groups associated with the lowest levels of self-reported expertise. The fact that these groups are very similar suggests that the forecasters are poor at assessing how little they know about a subject that is strange to them. If these groups are combined into a single group, the posterior probability of a strictly decreasing level of under-confidence is approximately 1.0.

@@

\section{Discussion}
This paper began with an introduction of a rather unorthodox but nonetheless realistic probability forecasting setting, in which forecasts are made very infrequently by a heterogeneous group of experts over a period of time. The resulting data is too sparse to be modeled efficiently with standard time-series methods.  In response to this lack of appropriate methodology, our work 
%expands the literature towards less-studied areas of  probability aggregation by
introduces an interpretable time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated crowd belief across time. The model is estimated in two-steps: The first step samples constrained versions of the model parameters via a Gibbs sampler that only makes use of standard distributions. The second step involves a one-dimensional optimization procedure that transforms the constrained parameter values into their unconstrained counterparts.
